Predicting complex diseases: performance, robustness and ..utility..(impact?)

### 2017-12-01 paper/draft.tex
save

Polygenic Risk Scores (PRSs) combine the information contained in many single-nucleotide polymorphisms (SNPs) in a score that aims at reflecting the risk of developing a disease for which this score has been created. PRSs have proven to be useful for different applications, such as finding a common genetic contribution between two diseases \cite[]{purcell2009common}. 
Personalized medicine is another major application of PRSs. Personalized medicine will use PRSs in screening campaigns in order to identify high-risk individuals for a given disease. As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}. Moreover, screening based on sequencing seems to make individuals positively change their health behavior while not causing patient anxiety nor depression \cite[]{vassy2017impact}.
Yet, PRSs would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases. For example, for respectively screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than respectively 75\% and 99\% \cite[]{janssens2007impact}.

@@

We recently developed two R packages, bigstatsr and bigsnpr, for efficient management and analysis of large-scale genome-wide data \cite[]{prive2017efficient}. Package bigstatsr includes efficient functions for computing penalized linear and logistic regressions on huge datasets. In this paper, we present a comprehensive comparative study of the P+T method, our penalized logistic regression and finally the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in \cite{botta2014exploiting}. 
Note that the SVM model is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp} and therefore isn't added to the comparison.
For the P+T model, we compare different thresholds of inclusion of SNPs.
For the logistic regression, we include two novel approaches. 
First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for choosing the amount of regularization used, which directly affects the number of SNPs included in the model. 
We also show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.

>>>>

Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRSs have proven to be useful for different applications, such as finding a common genetic contribution between two diseases \cite[]{purcell2009common}. 
Personalized medicine is another major application of PRSs. Personalized medicine will use PRSs in screening campaigns in order to identify high-risk individuals for a given disease. As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}. Moreover, screening based on sequencing seems to make individuals positively change their health behavior while not causing patient anxiety nor depression \cite[]{vassy2017impact}.
Yet, PRSs would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases. For example, for respectively screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than respectively 75\% and 99\% \cite[]{janssens2007impact}.

@@

We recently developed two R packages, bigstatsr and bigsnpr, for efficient management and analysis of large-scale genome-wide data \cite[]{prive2017efficient}. Package bigstatsr includes efficient functions for computing penalized linear and logistic regressions on huge datasets. In this paper, we present a comprehensive comparative study of the P+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in \cite{botta2014exploiting}. 
Note that the SVM model is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp} and therefore isn't added to the comparison.
For the P+T model, we compare different thresholds of inclusion of SNPs.
For the logistic regression, we include two novel approaches. 
First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for choosing the amount of regularization used, which directly affects the number of SNPs included in the model. 
We also show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.
